\bf Summary

### 2025-02-12 paper/paper.qmd
this version successfully submitted to arXiv NB needed manual intervention: - commenting out of line number funtionality (package and commands) - in text author citations (e.g. Weiss et al. (2020), rather than (Weiss et al. 2020) needed edits in `paper.tex` changing `\citet{authoryear}` to `Author et al. \citep{authoryear}`

---
title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  plos-pdf:
    keep-tex: true   
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliations:
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
      - Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliations:
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
      - Curtin University, Bentley, WA, Australia
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliations:
      - University of Western Australia, WA, Australia
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliations:
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
      - Curtin University, Bentley, WA, Australia
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations. We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [@tierney2018novel], setting rehabilitation districts for stroke patients [@padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].

>>>>

---
title: "`traveltime`: an R package to calculate travel time across a landscape from user-specified locations"
format:
  plos-pdf:
    keep-tex: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300   
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliations:
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
      - Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliations:
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
      - Curtin University, Bentley, WA, Australia
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliations:
      - University of Western Australia, WA, Australia
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliations:
      - The Kids Research Institute Australia, Nedlands 6009 WA, Australia
      - Curtin University, Bentley, WA, Australia
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations. We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias; @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [@tierney2018novel], setting rehabilitation districts for stroke patients [@padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].

### 2025-02-12 paper/paper.tex
this version successfully submitted to arXiv NB needed manual intervention: - commenting out of line number funtionality (package and commands) - in text author citations (e.g. Weiss et al. (2020), rather than (Weiss et al. 2020) needed edits in `paper.tex` changing `\citet{authoryear}` to `Author et al. \citep{authoryear}`

% line numbers
\usepackage[right]{lineno}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={traveltime: an R package to calculate walking and motorised travel time from user-specified locations},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\texttt{traveltime}: an R package to calculate walking
and motorised travel time from user-specified
locations} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
Gerard E. Ryan\textsuperscript{1,2*}, Nicholas
Tierney\textsuperscript{1,3}, Nick Golding\textsuperscript{4,1}, Daniel
J. Weiss\textsuperscript{1,3}
\\
\bigskip
\textbf{1} The Kids Research Institute Australia, Nedlands 6009 WA,
Australia, \\ \textbf{2} Melbourne School of Population and Global
Health, University of Melbourne, 3010, VIC,
Australia, \\ \textbf{3} Curtin University, Bentley, WA,
Australia, \\ \textbf{4} University of Western Australia, WA,
Australia, 
\bigskip

@@

\linenumbers

@@

Understanding and mapping the time to travel among locations is useful
for many activities from urban planning \citep{zahavi1974traveltime} to
public health \citep{hulland2019travel, weiss2020global} and myriad
others \citep{nelson2019suite}. Here we present a software package ---
\texttt{traveltime} --- written in and for the language R \citep{Rref}.
\texttt{traveltime} enables a user to create a map of the motorised or
walking travel time over an area of interest from a user-specified set
of geographic coordinates. The result is a raster of the area of
interest where the value in each cell is the lowest travel time in
minutes to any of the specified locations. We envisage this software
having diverse applications including: estimating sampling bias in
species occurrence data \citep[@reddy2003geographical]{dennis2000bias},
mapping electric vehicle charger accessibility
\citep{falchetta2021electric}, allocating public defibrillators
\citep{tierney2018novel}, setting rehabilitation districts for stroke
patients \citep{padgham2019introduction}, or understanding access to
agricultural processing facilities \citep{zhao2023replanting}.

@@

\texttt{traveltime} provides a spatial interface using object classes
from the \texttt{terra} package \citep{terra}. It accepts points as
\texttt{matrix}, \texttt{data.frame}, or \texttt{SpatVector} class
objects; and the area of interest as an extent in the form of a
\texttt{numeric}, \texttt{SpatExtent}, \texttt{SpatVector}, or
\texttt{SpatRaster} class object; and returns the result as a
\texttt{SpatRaster} class object. The travel time is calculated as
movement over a resistance `friction surface' \citep{gdistance2017}. To
provide easy access to the existing friction surfaces generated by
\citet{weiss2020global}, \texttt{traveltime} uses the package
\texttt{malariaAtlas} to allow users to access friction surfaces for the
area of interest; though users can also supply any other friction
surfaces to \texttt{traveltime}. Functionality from \texttt{gdistance}
\citep{gdistance2017} is also used internally to calculate the minimum
least-cost-distance for each cell from the points of interest.

@@

Global maps of travel time to cities
\citep{weiss2018global, nelson2019suite} and health care facilities
\citep{hulland2019travel, weiss2020global} have generated significant
interest and use\footnote{Collectively \textgreater1600 citations per
  Google Scholar at the 28th of January 2025.}, and the city data set of
\citet{nelson2019suite} is available to R users through the widely-used
\texttt{geodata} package \citep{geodata}. There is clear demand for
these type of products.

\citet{weiss2020global} made their code available as an R script to
allow for reproduction and extension of their analyses
(\url{https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt}
). To further enable extension of this work, here we have developed an R
package based on that code to seamlessly calculate the travel time from
any arbitrary set of locations.

@@

The \texttt{stations} data set included in the \texttt{traveltime}
package is a 563 row, 2 column \texttt{matrix} containing the longitude
(\texttt{x}) and latitude (\texttt{y}) of all LRT and MRT station exits
in Singapore from \citet{singdata}:

@@

Now that we have the two items of data that we require, the next step is
to prepare a friction surface for our area of interest. We will use the
friction surface from \citet{weiss2020global} that can be downloaded by
\texttt{traveltime} with the function \texttt{get\_friction\_surface()}.
We can pass in our basemap \texttt{singapore\_shapefile}, a
\texttt{SpatVector}, directly as the \texttt{extent}. We're interested
in walking time from a station, so we'll download the walking friction
surface by specifying \texttt{surface\ =\ "walk2020"}. (Alternatively,
we could use \texttt{surface\ =\ "motor2020"} for motorised travel if
that were of interest.) We're also only interested in walking \emph{on
land} so we then mask out areas outside of the land boundary in
\texttt{singapore\_shapefile}:

@@

\includegraphics{paper_files/figure-pdf/fig-data-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-result-1.pdf}

@@

\nolinenumbers
\renewcommand\refname{References}
  \bibliography{paper.bib}

>>>>

% line numbers
% \usepackage[right]{lineno}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={traveltime: an R package to calculate travel time across a landscape from user-specified locations},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\texttt{traveltime}: an R package to calculate travel
time across a landscape from user-specified
locations} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
Gerard E. Ryan\textsuperscript{1,2*}, Nicholas
Tierney\textsuperscript{1,3}, Nick Golding\textsuperscript{4,1}, Daniel
J. Weiss\textsuperscript{1,3}
\\
\bigskip
\textbf{1} The Kids Research Institute Australia, Nedlands 6009 WA,
Australia, \\ \textbf{2} Melbourne School of Population and Global
Health, University of Melbourne, 3010, VIC,
Australia, \\ \textbf{3} Curtin University, Bentley, WA,
Australia, \\ \textbf{4} University of Western Australia, WA,
Australia, 
\bigskip

@@

% \linenumbers

@@

Understanding and mapping the time to travel among locations is useful
for many activities from urban planning \citep{zahavi1974traveltime} to
public health \citep{hulland2019travel, weiss2020global} and myriad
others \citep{nelson2019suite}. Here we present a software package ---
\texttt{traveltime} --- written in and for the language R \citep{Rref}.
\texttt{traveltime} enables a user to create a map of the motorised or
walking travel time over an area of interest from a user-specified set
of geographic coordinates. The result is a raster of the area of
interest where the value in each cell is the lowest travel time in
minutes to any of the specified locations. We envisage this software
having diverse applications including: estimating sampling bias in
species occurrence data \citep{dennis2000bias, reddy2003geographical},
mapping electric vehicle charger accessibility
\citep{falchetta2021electric}, allocating public defibrillators
\citep{tierney2018novel}, setting rehabilitation districts for stroke
patients \citep{padgham2019introduction}, or understanding access to
agricultural processing facilities \citep{zhao2023replanting}.

@@

\texttt{traveltime} provides a spatial interface using object classes
from the \texttt{terra} package \citep{terra}. It accepts points as
\texttt{matrix}, \texttt{data.frame}, or \texttt{SpatVector} class
objects; and the area of interest as an extent in the form of a
\texttt{numeric}, \texttt{SpatExtent}, \texttt{SpatVector}, or
\texttt{SpatRaster} class object; and returns the result as a
\texttt{SpatRaster} class object. The travel time is calculated as
movement over a resistance `friction surface' \citep{gdistance2017}. To
provide easy access to the existing friction surfaces generated by
Weiss et al. \citep{weiss2020global}, \texttt{traveltime} uses the package
\texttt{malariaAtlas} to allow users to access friction surfaces for the
area of interest; though users can also supply any other friction
surfaces to \texttt{traveltime}. Functionality from \texttt{gdistance}
\citep{gdistance2017} is also used internally to calculate the minimum
least-cost-distance for each cell from the points of interest.

@@

Global maps of travel time to cities
\citep{weiss2018global, nelson2019suite} and health care facilities
\citep{hulland2019travel, weiss2020global} have generated significant
interest and use\footnote{Collectively \textgreater1600 citations per
  Google Scholar at the 28th of January 2025.}, and the city data set of
Nelson et al. \citep{nelson2019suite} is available to R users through the widely-used
\texttt{geodata} package \citep{geodata}. There is clear demand for
these type of products.

Weiss et al. \citep{weiss2020global} made their code available as an R script to
allow for reproduction and extension of their analyses
(\url{https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt}
). To further enable extension of this work, here we have developed an R
package based on that code to seamlessly calculate the travel time from
any arbitrary set of locations.

@@

The \texttt{stations} data set included in the \texttt{traveltime}
package is a 563 row, 2 column \texttt{matrix} containing the longitude
(\texttt{x}) and latitude (\texttt{y}) of all LRT and MRT station exits
in Singapore from Land Transport Authority \citep{singdata}:

@@

Now that we have the two items of data that we require, the next step is
to prepare a friction surface for our area of interest. We will use the
friction surface from Weiss et al. \citep{weiss2020global} that can be downloaded by
\texttt{traveltime} with the function \texttt{get\_friction\_surface()}.
We can pass in our basemap \texttt{singapore\_shapefile}, a
\texttt{SpatVector}, directly as the \texttt{extent}. We're interested
in walking time from a station, so we'll download the walking friction
surface by specifying \texttt{surface\ =\ "walk2020"}. (Alternatively,
we could use \texttt{surface\ =\ "motor2020"} for motorised travel if
that were of interest.) We're also only interested in walking \emph{on
land} so we then mask out areas outside of the land boundary in
\texttt{singapore\_shapefile}:

@@

\includegraphics{paper_files/figure-pdf/fig-data-1.png}

@@

\includegraphics{paper_files/figure-pdf/fig-result-1.png}

@@

% \nolinenumbers
\renewcommand\refname{References}
  \bibliography{paper.bib}
